Stateful Forward Reasoning

### 2025-02-04 paper.tex
Always end paragraph headings with periods

\paragraph{ACL2}
Among white-box proof automation tactics, only ACL2's \enquote{waterfall} (TODO ref) has special support for forward reasoning.
(Some such tactics, e.g.\ Coq's \texttt{auto}, can execute arbitrary tactics including forward rules, but make no special provisions for them.)
Like Aesop, ACL2 works on goals and saturates the local context of each goal, using user-provided forward rules, before using backward reasoning proof methods.
It uses essentially Aesop's naive algorithm, i.e.\ the local context of each goal is saturated independently without reusing partial results from other goals.

\paragraph{SATCHMO}
The Prolog-based first-order theorem prover SATCHMO (TODO ref) and its many descendants (TODO refs) uses a tableau calculus (TODO ref) that can be viewed as combined forward and backward reasoning.
Given a set of implications $A₁ ∧ \dots ∧ Aₙ → B₁ ∨ \dots ∨ Bₘ$ (with special cases $⊤ → B₁ ∨ \dots ∨ Bₘ$ and $A₁ ∧ \dots ∧ Aₙ → ⊥$), SATCHMO attempts to create a model that satisfies all implications.
To that end, it matches the premises $A₁, \dots, Aₙ$ of each rule against known facts from Prolog's database, which is initially empty.
If this is successful, SATCHMO first adds $B₁$ to the Prolog database and continues its search; if the continuation fails (i.e., if $⊥$ is derived), SATCHMO backtracks and tries $B₂$, etc.

@@

\paragraph{Knowledge-based reasoning systems}
Combined forward and backward reasoning is also supported by some knowledge-based reasoning systems, e.g.\ Algernon (TODO ref) and EYE (TODO ref).
However, to our knowledge none of these systems have backwards rules that can change the facts available to forward rules.
Hence backward and forward reasoning do not interact and there is no need for our technique.
Nevertheless, our proposal bears some resemblance to the RETE algorithm (TODO ref) for forward reasoning in large knowledge bases.

\paragraph{Other related work}
More distantly related are various calculi that combine backward reasoning with some form of incremental forward reasoning or simplification.
Examples include the DPLL algorithm with its unit resolution rule (TODO ref) or incremental theory reasoners in SMT (TODO ref) and tableaux (TODO refs).
Focusing (TODO ref) provides a proof-theoretic view on the interplay of forward and backward reasoning.

>>>>

\paragraph{ACL2.}
Among white-box proof automation tactics, only ACL2's \enquote{waterfall} (TODO ref) has special support for forward reasoning.
(Some such tactics, e.g.\ Coq's \texttt{auto}, can execute arbitrary tactics including forward rules, but make no special provisions for them.)
Like Aesop, ACL2 works on goals and saturates the local context of each goal, using user-provided forward rules, before using backward reasoning proof methods.
It uses essentially Aesop's naive algorithm, i.e.\ the local context of each goal is saturated independently without reusing partial results from other goals.

\paragraph{SATCHMO.}
The Prolog-based first-order theorem prover SATCHMO (TODO ref) and its many descendants (TODO refs) uses a tableau calculus (TODO ref) that can be viewed as combined forward and backward reasoning.
Given a set of implications $A₁ ∧ \dots ∧ Aₙ → B₁ ∨ \dots ∨ Bₘ$ (with special cases $⊤ → B₁ ∨ \dots ∨ Bₘ$ and $A₁ ∧ \dots ∧ Aₙ → ⊥$), SATCHMO attempts to create a model that satisfies all implications.
To that end, it matches the premises $A₁, \dots, Aₙ$ of each rule against known facts from Prolog's database, which is initially empty.
If this is successful, SATCHMO first adds $B₁$ to the Prolog database and continues its search; if the continuation fails (i.e., if $⊥$ is derived), SATCHMO backtracks and tries $B₂$, etc.

@@

\paragraph{Knowledge-based reasoning systems.}
Combined forward and backward reasoning is also supported by some knowledge-based reasoning systems, e.g.\ Algernon (TODO ref) and EYE (TODO ref).
However, to our knowledge none of these systems have backwards rules that can change the facts available to forward rules.
Hence backward and forward reasoning do not interact and there is no need for our technique.
Nevertheless, our proposal bears some resemblance to the RETE algorithm (TODO ref) for forward reasoning in large knowledge bases.

\paragraph{Other related work.}
More distantly related are various calculi that combine backward reasoning with some form of incremental forward reasoning or simplification.
Examples include the DPLL algorithm with its unit resolution rule (TODO ref) or incremental theory reasoners in SMT (TODO ref) and tableaux (TODO refs).
Focusing (TODO ref) provides a proof-theoretic view on the interplay of forward and backward reasoning.
